# Chapter. Logistic regression Analysis

```{r}
date()
```
**About the Data set**

*The data was collected for the following paper:*
*P. Cortez and A. Silva. Using Data Mining to Predict Secondary School Student Performance. In A. Brito and J. Teixeira Eds., Proceedings of 5th FUture BUsiness TEChnology Conference (FUBUTEC 2008) pp. 5-12, Porto, Portugal, April, 2008, EUROSIS, ISBN 978-9077381-39-7.*

*The secondary student achievement of two Portuguese schools is the focus of these data. The data variables were gathered utilizing school reports and surveys and include student grades, demographic, social, and school-related features. Regarding the performance in two different courses, Mathematics (mat) and Portuguese language(por), two datasets are supplied . The two datasets were modeled in [Cortez and Silva, 2008] under binary/five-level classification and regression tasks. Please take note that the target attribute G3 strongly correlates with the traits G2 and G1. This is due to the fact that G3 is the final year grade (given at the third period), but G1 and G2 are the grades for the first and second periods, respectively.*

## Reading data

```{r}
setwd("C:/Users/Aadhar/Desktop/Uni/Open Data Science/data")

library(readr)
alc <- read_csv("C:/Users/Aadhar/Desktop/Uni/Open Data Science/data/alc.csv",show_col_types = FALSE)
alc
```
**Observation**
*The data set, named here as "alc", has 370 observations of 35 variables*

## Visualising data through plots

```{r}
# access the tidyverse libraries tidyr, dplyr, ggplot2
library(tidyr); library(dplyr); library(ggplot2)

# glimpse at the alc data
glimpse(alc)

# use gather() to gather columns into key-value pairs and then glimpse() at the resulting data
gather(alc) %>% glimpse

# it may help to take a closer look by View() and browse the data
gather(alc) %>% View

# draw a bar plot of each variable
gather(alc) %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```
*Above is the plot of each variable in the data set and represent how each variable is distributed*

## About my 4 chosen variables

*I am choosing the following variable*
**schoolsup** *- extra educational support(binary: yes or no)*
**higher** *- wants to take higher education (binary: yes or no)*
**famrel** *- quality of family relationships (numeric: from 1 - very bad to 5 - excellent)*
**final grade** *numeric: from 0 to 20, output target*

*The variable 'schoolsup' will help me to understand the impact of high alcohol use on how much support the student need in their studies.*
**My hypothesis** *is that with increased alcohol use, there will be increase the support needed*

*With the help of the variable 'higher', I want to study whether the willingness to pursue higher studies changes between those who consume more alcohol than others.*
**My hypothesis** *is that there will be an effect but not so significant on desire to pursue higher education with increased alcohol consumption*

*the 'famrel' variable will help me understand the effect of high alcohol usage on the relationship with family.*
**My hypothesis** *is that there will be a strain in relationship with family with high consumption*

*'final grade' variable with help me see if there is change in final grades of the student with higher usage as compared to lower usage*
**My hypothesis** *is that the marks will be negatively effected with high consumption of alcohol*

#Exploring the chosen variables through visualization

#1.Exploring high_use with schoolsup i.e., extra support need from school

```{r}

# access the tidyverse libraries dplyr and ggplot2
library(dplyr); library(ggplot2)

g1 <- ggplot(alc, aes(x = high_use))

g1 + geom_bar() + facet_wrap("schoolsup") + ggtitle("Effect of alcohol consumption on Educational support required") + xlab("High alcohol usage") + aes(fill = high_use)
```

**Observation**
*From the above visualization we can observe that, there are comparatively less students whose alcohol usage is high and requires extra educational support. Interestingly, students for those alcohol usage isn't high, they seek more support from the school when compared to those with high usage.*
*My hypothesis stands false for this which is very interesting for me*

#2. Exploring high_use with higher i.e. desire for higher eduation

```{r}
g2 <- ggplot(alc, aes(x = high_use))

g2 + geom_bar() + facet_wrap("higher") + ggtitle("Desire to take higher education vs high alcohol usage") + xlab("High alcohol usage") + aes(fill = high_use)
```

**Observation**
*The above variable stands true to my hypothesis that no matter how much is the consumption for alcohol, the desire for higher education doesn't go away. Lets take a closer look below*

#Summarising to see results see more clearly

```{r}
alc %>% group_by(higher, high_use) %>% summarise(count = n())
```

**Observation**
*There are only 16 students that do not desire to pursue higher education out of which 9 were with high usage. This means that there is no significant relationship between alcohol usage and desire to study further, which seems logical to me as food choices can be temporary but the choice for further studies is a thought through decision for most people.*

#3. Exploring high_use and its effect on famrel i.e., family relationship

```{r}
g3 <- ggplot(alc, aes(x = high_use))

g3 + geom_bar() + facet_wrap("famrel") + ggtitle("family relationships (1-5) vs high alcohol usage") + xlab("High alcohol usage") + aes(fill = high_use)
```

**Observation**
*Family relationship for most students have been on the positive side without any significant change with high or low usage of alcohol. Infact in the category of bad relationship, the proportion of those with low usage is higher but that is true as there are more number of students with low usage as compared to high usage in general.*
*Therefore, my hypothesis of effect being negative stands false*

#4. Exploring high_use and its effect on G3 i,e., Final Grades of student

```{r}
g4 <- ggplot(alc, aes(x = high_use, y = G3))

g4 + geom_boxplot() + ggtitle("Effect of high alcohol usage on Final grades") + xlab("High alcohol usage") + ylab("Final grade")+ aes(col = high_use)
```
**Observation**
*We can clearly visualize that the students with high consumption have comparatively lower grades than those with low consumption.*
*The hypothesis for the variable stands true that high alcohol consumption has negatively effected the grades of students*

## Logistic Regression Model

**About Logistic regression model**

*Logistic regression is used to predict the class (or category) based on one or multiple predictor variables (x). It is used to model a binary outcome, that is a variable, which can have only two possible values*

*We will now use Logistic regression model to identify factors related to higher than average student alcohol consumption.*

```{r}
# find the model with glm()
m <- glm(high_use ~ schoolsup + higher + famrel + G3, data = alc, family = "binomial")

# print out a summary of the model
summary(m)

# print out the coefficients of the model
coef(m)
```
**Observation**
*We can see that p-value for all the variables, is less than alpha(0.5) which means that with increase in one unit of alcohol consumption, there is a significant rise in support need from school and subsequently similar result can be seen for other variables i.e., there is significant rise in 'higheryes', 'famrel', and 'G3' respectively* 

*There we 4 Fisher scoring iterations before the process of fitting the model stopped and output the results.*

*In this case, our intercept for different variable is less than 0, it matches with our p-value results observation being less than the alpha*


## From coefficients to odds ratios(OR)

```{r}
# find the model with glm()
m <- glm(high_use ~ schoolsup + higher + famrel + G3, data = alc, family = "binomial")

# compute odds ratios (OR)
OR <- coef(m) %>% exp

# compute confidence intervals (CI)
CI <- confint(m) %>% exp

# print out the odds ratios with their confidence intervals
cbind(OR, CI)
```
**Observation**
*The OR value suggest that when one unit of alcohol consumption (high_use) increases, the odds of educational support decreases 21% and with one unit increase in schoolsup, odds of higher education decreases by 55% and odds of good fam decreases by 24% and by 8% for grades*
*The Odds ratio of The final grade is 1:1:0:0*

#Plot Visualisation by providing Confidence Intervals
```{r}
library(finalfit)
dependent <- "high_use"
explanatory <- c("schoolsup", "higher", "famrel", "G3")
alc %>% 
  or_plot(dependent, explanatory,
          breaks = c(0.5, 1, 2, 5),
          table_text_size = 3.5,
          title_text_size = 16)
```

*The results of the above plot matches the observations*

## Explore the predictive power of the model with the significant variables

#Binary prediction (1)

```{r}
# fit the model
m <- glm(high_use ~ schoolsup + higher + famrel + G3, data = alc, family = "binomial")
m1 <- glm(high_use ~ famrel + G3, data = alc, family = "binomial")

# predict() the probability of high_use
probabilities <- predict(m, type = "response")
prob2 <- predict(m1, type = "response")

library(dplyr)

# add the predicted probabilities to 'alc'
alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prob2 = prob2)

alc <- mutate(alc, prediction = probability > 0.5)

# use the probabilities to make a prediction of high_use
alc <- mutate(alc, prob3 = prob2 > 0.5)

# see the last ten original classes, predicted probabilities, and class predictions
select(alc, schoolsup, higher, famrel, G3, high_use, probability, prediction) %>% tail(10)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction)

```
**Observation**
*The model has predicted False as 251/259*
*The model has predicted True as 8/111*
*The sensitivity of the model is 8 which is not so good and specificity is 251 which is much better*

*In the cross table, 8 times the prediction is heavy alcohol consumption when the variable gets a value corresponding to no heavy alcohol consumption. Similarly, 103 times the prediction is no high alcohol consumption when the variable gets a value corresponding to high alcohol consumption.*

```{r}
# rerunning some codes to avoid errors

m <- glm(high_use ~ schoolsup + higher + famrel + G3, data = alc, family = "binomial")
alc <- mutate(alc, probability = predict(m, type = "response"))
alc <- mutate(alc, prediction = probability > 0.5)
```


#Binary prediction (2)

```{r}

# initialize a plot of 'high_use' versus 'probability' in 'alc'
g <- ggplot(alc, aes(x = probability, y = high_use))

# define the geom as points and draw the plot
g + geom_point() + aes(col = probabilities)

# tabulate the target variable versus the predictions
table(high_use = alc$high_use, prediction = alc$prediction) %>% prop.table() %>% addmargins()

```
**Observation**
*The table above shows that, according to the forecast, there are about 95 percent of all people who do not drink a lot of alcohol. There are approximately 70 percent of all persons who do not drink much alcohol. Note again that the forecast differs from the values of the variable.*

## Accuracy and loss functions

```{r}
# define a loss function (mean prediction error)
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}

# call loss_func to compute the average number of wrong predictions in the (training) data
loss_func(class = alc$high_use, alc$probability)

```
**Observation**
*The loss function gave the prediction error of 0.3*

## K-fold Cross validation
```{r}

# K-fold cross-validation
library(boot)

cv <- cv.glm(data = alc, cost = loss_func, glmfit = m, K = 10)


# average number of wrong predictions in the cross validation
cv$delta[1]
```
**Observation**
*Since the lower value means better this gives a comparatively poor result than the exercise result which was 0.26*

**____End of Chapter___**
